# Helper functions

## Introduction

In addition to the main **dplyr** verbs, the package provides quite a few **helper functions**. Helper functions are used in conjunction with the main verbs to make specific tasks and calculations a bit easier. Many of these are summarised in the **dplyr** [cheat sheat](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) (under *Manipulate Variables*, *Vector Functions* and *Summary Functions*). We are not going to review every single one of them in this chapter. Instead, we aim to point out where helper functions tend to be used and highlight a few of the more useful ones.

## Working with `select`

There are a few helper functions that can be used with `select`. Their job is to make it easier to match variable names according to various criteria. We'll look at the three simplest of these---look at the examples in the help file for `select` and the [cheat sheat](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) to see what else is available.

We can select variables according to the sequence of characters used at the start of their name with the `starts_with` function. For example, to select all the variables in `penguins` that begin with the word "bill", we use:

```{r}
select(penguins, starts_with("Bill"))
```

This returns a tibble containing just `bill_length_mm` and `bill_depth_mm`. There is also a helper function to select variables according to characters used at the end of their name---the `ends_with` function (no surprises there).

Notice that we quote the name we want to match against because `starts_with` expects a literal character value. This is not optional. Unusually, `starts_with` and `ends_with` are not case sensitive by default. For example, we passed `starts_with` the argument `"Bill"` instead of `"bill"`, yet it still selected variables beginning with the character string `"bill"`. If we want to select variables on a case-sensitive basis, we need to set an argument `ignore.case` to `FALSE` in `starts_with` and `ends_with`.

The last `select` helper function we'll look at is called `contains`. This one allows us to select variables based on a partial match **anywhere** in their name. Look at what happens if we pass `contains` the argument `"length"`:

```{r}
select(penguins, contains("length"))
```

This selects all the variables with the word 'length' in their name.

There is nothing to stop us combining the different variable selection methods. For example, we can use this approach to select all the variables that have some units at the end of their names (millimetres or grams):

```{r}
select(penguins, ends_with("_mm"), ends_with("_g"))
```

When we apply more than one selection criteria like this, the `select` function returns the variables that match any criteria, rather than the set that meets all of them.

## Working with `mutate` and `transmute`

There are quite a few **helper functions** that can be used with `mutate`. These make it easier to carry out certain calculations that aren't easy to do with base R. We won't explore these here as they tend to be needed only in quite specific circumstances. However, in situations where we need to construct an unusual variable, it's worth looking at that [handy cheat sheat](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) to see what options might be available.

We will look at one particularly useful helper function that's used a lot when we need to recode a particular variable using `mutate`. The function is called `case_when`. It works by setting up a series of paired matching criteria and replacement values. For example, imagine that we want to replace the names in `species` with three-letter shortcodes for each species. This is how to achieve that using `case_when` with `mutate`:

```{r}
penguins %>% 
  mutate(species = case_when(
           species == "Adelie"    ~ "ADL",
           species == "Gentoo"    ~ "GEN",
           species == "Chinstrap" ~ "CHN",
           TRUE ~ "UNKNOWN"
         )) 
```

The `mutate` bit of this is not new. Look at the `case_when` component---there are four criteria. The first of these is `species == "Adelie"    ~ "ADL"` . The way to read this is, "look for cases where the value of `species` is equal to `"Adelie"`, and where that is true spit out the value `"ADL"`". `case_when` steps through each criterion like this in turn, trying to find a match. The last one `TRUE ~ "UNKNOWN"` acts as a catch-all for the non-matches.

This looks confusing at first but it does make sense with a bit of practise, and recoding variables using `case_when` is a lot easier than going through a spreadsheet by hand.

## Working with `filter`

There aren't that many **dplyr** helper function that works with `filter`. In fact, we've already looked at the most useful one: the `between` function. This is used to identify cases where the values of a numeric variable lie inside a defined range. For example, if we want all the individuals that had a body mass in the 4-5kg range, we could use:

```{r}
filter(penguins, between(body_mass_g, 4000, 5000))
```

## Working with `summarise`

There are a small number **dplyr** helper functions that can be used with `summarise`. These generally provide summaries that aren't available directly using base R functions. For example, we've already seen the `n_distinct` function in action. This can be used to calculate the number of distinct values of a variable:

```{r}
summarise(penguins, 
          num_species = n_distinct(species),
          num_island  = n_distinct(island ))
```

This confirms what we already knew---that there are three unique species and three unique islands in the `penguins` data set.
