```{r, echo=FALSE}
opts_chunk$set(warning=FALSE)
```

# Customising plots {#chapter-customise-ggplot2}

The default formatting used by **ggplot2** has been carefully chosen to ensure the information in a plot is easy to discern. For example, by default, plots include a pale grey background and include grid lines. This is designed to emphasise the data while supporting comparisons. The grey highlights colour differences and ensures the grid lines have little visual impact beyond aiding in the assessment of position. Perfect for exploratory analysis!

Although we can justify these sorts of choices on the grounds they improve the readability of a plot, the fact remains they are somewhat unconventional and not much-loved by many users. For this reason, we often need to change the appearance of a plot before we include it in a document or presentation---e.g. most published figures use a white, rather than grey, background.

This chapter will demonstrate how to customise **ggplot2** plots. We are not going to cover every possible permutation. That would need its [own book](https://ggplot2-book.org/). Instead, we will explore the main routes to customisation so that we can use these as we review different visualisations later. Using the `penguins` data once again, we'll work on improving the following scatter plot from the [Introduction to **ggplot2**](#chapter-ggplot2-intro) chapter:

```{r, fig.width=4.5}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer
  geom_point()
```

## Geom properties

One common way to change a plot's appearance is by altering the properties of one or more geometric objects, such as points, lines, bars or polygons. How do we change the properties of the geom associated with a particular layer? We set those properties by specifying the appropriate arguments in the defining `geom_NAME` function.

For example, how might we change properties like the size or shape of points in our scatter plot? We used the `geom_point` function to add points to the scatter plot, which means we have to set the arguments of `geom_point` to change the point properties. We can use this idea to rebuild the example scatter plot, setting the shape, size and transparency of points:

```{r, fig.width=4.5}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85)
```

The point shape is set with the `shape` argument. There are a few different ways to specify point shapes in R. We used the numeric coding system in this example (17 = triangle). Unfortunately, the numeric codes aren't at all intuitive. Here are the more common ones:

```{r, echo=FALSE, fig.width=7, fig.height=0.5}
shapes <- data.frame(p = 0:25); ggplot() + scale_shape_identity() + geom_point(data=shapes, mapping=aes(x=p, y=1, shape=p), size=5, fill="red") + geom_text(data=shapes, mapping=aes(x=p, y=0, label=p), size=4) + ylim(-0.5, 1.5) + theme_void()
```

The point size is specified by the `size` argument. This has a baseline default value of 1. We assigned `size` a value of 1.5, thereby increasing the point size relative to the default. We could make the points smaller by using a value less than 1.

We made the points semi-transparent by setting the value of the `alpha` argument to be less than 1. In graphical systems, the 'alpha channel' specifies transparency---a value of 0 is taken to mean 'completely invisible', and a value of 1 means 'completely opaque'.

There are other arguments---such as `fill` and `colour`---that can also be used to adjust the way the points are rendered. We'll look at these later.

The key message to take away from this customisation example is this---if we need to alter the properties of a geometric object (points, lines, ect), we do so by specifying the appropriate arguments in the `geom_NAME` function that defines the layer it belongs to.

### Relationship between aesthetic mappings and geom properties

In the previous chapter, we saw that we could introduce information into a plot by setting aesthetic mappings in the `aes` part of its specification. In the scatter plot example, we included information about species identity by mapping the `species` variable to the colour aesthetic. What happens if we also try to set the colour within the `geom_point` part?

```{r, fig.width=3.7}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour) + sex (shape)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points (including colour)
  geom_point(colour = "skyblue", shape = 17, size = 1.5, alpha = 0.85)
```

All the points are now one colour ('sky blue') which means we have lost the species information. What does this example demonstrate? When we set the properties of a geom, this will override any aesthetic mappings that conflict with our choice of customisation. Try to remember that---failing to do so is a good way to become frustrated.

::: {.infobox .information data-latex="{information}"}
#### Built-in colours in R {.unnumbered}

There is nothing special about 'skyblue' other than the fact that it is colour name known to R. There are over 650 built-in colour names. To see them, use a function called `colours` to print them all to the Console. Here are 25 selected at random:

```{r, echo=FALSE}
set.seed(27081975)
```

```{r}
sample(colours(), size = 25, replace = FALSE)
```
:::

## Plot scales

Setting the arguments of a `geom_` function applies changes in a layer-specific manner. Other kinds of customisation affect every layer in a plot. Remember what we said about **scales** in the previous chapter:

> The scale part of a **ggplot2** object controls how the information in a variable is mapped to the aesthetic properties. A scale takes the data and converts it into variation we can perceive, such as an x/y location or the colour and size of points in a plot.

Every aesthetic mapping has a scale associated with it. We adjust 'how the information in a variable is mapped to the aesthetic properties' by changing its corresponding scale.

As always, this kind of thing is best understood by example. Let's adjust the scale associated with the y-axis aesthetic by increasing the number of the horizontal lines and their accompanying labels (the 'guides').

```{r, fig.width=4.5}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  # customise y-axis grid line and label locations
  scale_y_continuous(breaks = seq(14, 20, by = 2))
```

What's going on here?

-   The aesthetic we wanted to alter was the y-axis position. The y-axis in this particular plot is associated with a continuous scale because `bill_depth_mm` is a numeric variable. This means we have to apply the `scale_y_continuous` function to tweak the position of the y-axis guides.

-   The `breaks` argument of `scale_y_continuous` takes a numeric vector to specify where the guides should be drawn. We used the base R `seq` function to set up a numeric vector containing the sequence: 14, 16, 18 and 20. That tells **gglot2** to place the guides at 14, 16, 18 and 20 on the y-axis.

The functions that adjust a scale all have the general form `scale_AES_TYP`, where:

-   the `AES` bit in the name refers to the relevant aesthetic, and
-   the `TYP` part refers to the type of scale we want to apply.

Each kind of `scale_AES_TYP` function has its own set of arguments to control how the scale works. For example, the `breaks` argument of `scale_y_continuous` uses a numeric vector to specify where the guides should be drawn. We can do much fancier things if we want to. For example, the `trans` argument can be used to apply a particular transformation to an axis, such as a logarithm.

Predictably, there is also a `scale_x_continuous` function. We can use this in exactly the same way to control where the guides on the x-axis appear:

```{r, fig.width=4.5}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  # customise x- and y-axis grid line and label locations
  scale_x_continuous(breaks = seq(34, 62, by = 4)) +
  scale_y_continuous(breaks = seq(14, 20, by = 2))
```

The key message to take away from this is that every aesthetic mapping has a scale associated with it. If we want to change how the information associated with an aesthetic mapping is displayed we have to change the corresponding scale. That suggests we can change the way point colours are associated with the `species` variable in our example by using one of the `scale_colour_YY` functions. That is correct:

```{r, fig.width=4.7}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  # customise x- and y-axis grid line and label locations
  scale_x_continuous(breaks = seq(34, 62, by = 4)) +
  scale_y_continuous(breaks = seq(14, 20, by = 2)) + 
  # customise the species colours ('value') and labels
  scale_color_manual(
    limits = c("Chinstrap", "Gentoo", "Adelie"),
    values = c("cornflowerblue", "seagreen", "orangered3"),
    labels = c(Chinstrap = "P. antarcticus", Gentoo = "P. papua", Adelie = "P. adeliae")
  )
```

This uses `scale_color_manual` to achieve three things:

-   it defines the order to display the three species in the legend via `limits`,
-   it sets up a custom colour palette via `values`, and
-   it alters the category labels via `labels` (so that their scientific names are used).

Scales are one of the hardest aspects of **ggplot2** to master. There are many different scales, each controlled by its own unique set of arguments. Fortunately, the defaults used by **ggplot2** are often good enough to arrive at a reasonable plot without having to manipulate the scales.

## Labels

What else might we like to tweak? Look at the x and y-axis labels. These are the names of the variables used in the x/y aesthetic mappings. These labels aren't too bad, but they could be nicer to look at. We know 'bill_depth_mm' stands for 'bill depth (measured in mm)', so why not use an axis label more like that?

The axes labels are a feature of the whole plot---i.e. they do not belong to a particular layer. This means we can't alter axis labels by passing arguments to the function that built a layer (`geom_point`). We have to use the labelling function: `labs`. This sets the label of any mapped aesthetics. For example, to set the labels of the x and y axes and the label associated with the colour mapping (i.e. the `species` legend title), use:

```{r, fig.width=4.7}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  # customise x- and y-axis grid line and label locations
  scale_x_continuous(breaks = seq(34, 62, by = 4)) +
  scale_y_continuous(breaks = seq(14, 20, by = 2)) + 
  # customise the species colours ('value') and labels
  scale_color_manual(
    limits = c("Chinstrap", "Gentoo", "Adelie"),
    values = c("cornflowerblue", "seagreen", "orangered3"),
    labels = c(Chinstrap = "P. antarcticus", Gentoo = "P. papua", Adelie = "P. adeliae")
  ) +
  # set the the x/y-axis labels and legend label using `labs`
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", colour = "Penguin Species")
```

Using `labs` is easy---if a mapping was set up in `aes` we can change its label via a 'name = value' construct in `labs`. That's it.

Finally, another way to set the axis labels is via the `xlab` and `ylab`. We tend not to use these because `labs` deals with all our labelling needs on its own. Why waste effort remembering three things if you only need to remember one?

## Themes

The final route to customisation happens via the 'theme' of a plot. We haven't considered the **ggplot2** theme system at all yet. A **ggplot2** theme deals with all the visual elements of a plot that aren't directly handled by adjusting geom properties or scales. These are essentially the 'non-data' parts of a plot---features such as the colour of the plotting region and the grid lines, whether or not those grid lines are even displayed, the position of labels, the font used in labels, and so on.

Here's a short example that adjusts the legend of the example plot:

```{r, fig.width=4.0, fig.asp=1.1}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  # customise x- and y-axis grid line and label locations
  scale_x_continuous(breaks = seq(34, 62, by = 4)) +
  scale_y_continuous(breaks = seq(14, 20, by = 2)) + 
  # customise the species colours ('value') and labels
  scale_color_manual(
    limits = c("Chinstrap", "Gentoo", "Adelie"),
    values = c("cornflowerblue", "seagreen", "orangered3"),
    labels = c(Chinstrap = "P. antarcticus", Gentoo = "P. papua", Adelie = "P. adeliae")
  ) +
  # use `labs` to set x/y-axis labels and remove legend label
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", colour = "") +
  # use theme to customise position and text formatting of the legend
  theme(
    legend.position = "top",
    legend.text = element_text(face="italic")
  )
```

Changes to the theme are specified using... the `theme` function. We used `theme` to move the legend to the top of the plot (via `legend.position`) and show the species labels in italics (via `legend.text`). We also got rid of the legend title by setting the colour label to empty character string in `labs`. We could do that via `theme` but using `labs` is simpler.

The **ggplot2** theme system is extremely powerful. Once we know how to use it, we can set up a custom theme and apply it with very little effort. However, that's not an entirely trivial thing to do because there are just so many components of every plot. Most people just google whatever adjustment they want to make when the time comes. Even then, tweaking each individual part of a theme can be very time consuming. Fortunately, there are a range of standard themes built into **ggplot2**. Here's how to apply a built-in theme:

```{r, fig.width=4.0, fig.asp=1.1}
# display bill morphology scatter plot
ggplot(penguins, 
       # bill depth (y) vs bill length (x) by species (colour)
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  # add points layer and customise appearance of points
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  # customise x- and y-axis grid line and label locations
  scale_x_continuous(breaks = seq(34, 62, by = 4)) +
  scale_y_continuous(breaks = seq(14, 20, by = 2)) + 
  # customise the species colours ('value') and labels
  scale_color_manual(
    values = c("cornflowerblue", "seagreen", "orangered3"),
    labels = c(Adelie = "P. adeliae", Chinstrap = "P. antarcticus", Gentoo = "P. papua")
  ) +
  # use `labs` to set x/y-axis labels and remove legend label
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", colour = "") +
  # apply 'black and white' theme
  theme_bw(base_size = 13) +
  # use theme to customise position and text formatting of the legend
  theme(
    legend.position = "top",
    legend.text = element_text(face="italic")
  )
```

This uses `+` with the `theme_bw` function to add the built-in 'black and white' theme. This theme dispenses with the grey background that so many people dislike. There are a couple of things to note from this example:

-   To apply the changes to the legend position, we had to place the `theme` part **after** we set the overall theme using `theme_bw`. The `theme_bw` function will override the `theme` changes if we try to do it the other way round.

-   We set the `base_size` argument of `theme_bw`. This rescales the text in a plot, leaving the relative size of the different elements unchanged. We did not have to do this, but it is worth showing off because it is such a useful trick.

There aren't that many themes built into **ggplot2**. Here's a quick look at them:

```{r, echo=FALSE, fig.show="hold", out.width="33.3%", fig.align='default'}
plt <- ggplot(penguins,
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  geom_point(shape = 17, size = 1.5, alpha = 0.85) + 
  scale_x_continuous(breaks = seq(34, 62, by = 4)) +
  scale_y_continuous(breaks = seq(14, 20, by = 2)) + 
  scale_color_manual(
    values = c("cornflowerblue", "seagreen", "orangered3"),
    labels = c(Adelie = "P. adeliae", Chinstrap = "P. antarcticus", Gentoo = "P. papua")
  ) + 
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", colour = "") 

my_theme <- 
  theme(legend.position = "none",
        plot.title = element_text(family = "Courier"))

plt + theme_grey   (base_size = 14) + ggtitle("theme_grey")    + my_theme
plt + theme_bw     (base_size = 14) + ggtitle("theme_bw")      + my_theme
plt + theme_minimal(base_size = 14) + ggtitle("theme_minimal") + my_theme
plt + theme_classic(base_size = 14) + ggtitle("theme_classic") + my_theme
plt + theme_light  (base_size = 14) + ggtitle("theme_light")   + my_theme
plt + theme_dark   (base_size = 14) + ggtitle("theme_dark")    + my_theme
```

<!--# remember to switch warnings on again -->

```{r, echo=FALSE}
opts_chunk$set(warning=TRUE)
```
