```{r, echo=FALSE}
opts_chunk$set(warning=FALSE)
```

# Exploring one variable

This chapter will consider how to go about exploring a single variable. Using the `penguins` data, we'll review some basic visualisations for exploring the sample distribution of numeric and categorical variables.

## Exploring numerical variables {#explore-numeric}

We'll work with flipper length (`flipper_length_mm`) and body mass (`body_mass_g`) to demonstrate how to explore numeric variables.

### What kind of numeric variable?

The simplest way to explore data is to view it in its raw form. Take a quick look at the first 100 values of `flipper_length_mm` and `body_mass_g`. We can get these by extracting them with the `$` operator and using the `head` function:

**flipper length** ---

```{r}
head(penguins$flipper_length_mm, 50)
```

**body mass** ---

```{r}
head(penguins$body_mass_g, 50)
```

These are clearly both numeric variables, but we can say a bit more:

-   They are both measured on a ratio scale. Zero really is zero, meaning that it makes sense to say that a length of 200 millimetres is twice as long as 100 millimetres or that 6000 grams is twice the mass of 3000 grams.
-   They are continuous variables. Think about the possible values flipper length and body mass can take. A length of 200.52 millimetres and a mass of 3000.71 grams are both perfectly reasonable, so fundamentally, these are continuous variables.

Notice that even though body mass is a continuous variable `body_mass_g` looks discrete because body mass measurements have been taken to the nearest 25 grams. This reflects methodological limitations---presumably, weighing a live penguin to more than 25 grams precision in the field is challenging. This illustrates an important idea: we can't just look at the values a numeric variable takes to determine whether it is discrete or continuous.

Whether we treat a variable as continuous or discrete is sometimes an analysis decision. We have to decide based on knowledge of a variable's true nature and the measurement process. For example, if we were only able to measure body mass to the nearest 500 grams, we would only "see" a few different body mass categories. It might then be reasonable to treat the `body_mass_g` variable as an ordinal, categorical variable.

### Histograms

We only looked at the first 50 values of the `flipper_length_mm` and `body_mass_g` variables because the `penguins` data set is a bit too big to view everything at once easily. It's hard to say much about the sample distribution of a numeric variable by looking at the raw data because there are so many unique values. Those first 50 are also not representative of the whole sample.

What else might we do? One useful tool is 'binning'. The idea behind binning a numeric variable is very simple. It involves two steps.

1.  Divide the number line into equal-sized, non-overlapping intervals. These are the 'bins'. We pick the interval size to ensure bins (typically) include at least a few cases.
2.  We then have to count up the number of times the variable falls inside each bin. The resulting set of counts summarises the sample distribution of the numeric variable.

Binning is very tedious to do by hand. Fortunately, R can do this for us. This shows how to use the `cut` function with the **dplyr** `transmute` and `count` verbs to bin the `flipper_length_mm` variable into 5 mm intervals:

```{r}
penguins %>% 
  transmute(bins = cut(flipper_length_mm, breaks = seq(170, 235, by = 5))) %>%
  count(bins)
```

We won't explain how this works because we only need to understand the output. This is a small tibble where the `bins` column shows the intervals, and the `n` column gives the associated counts.

What does this tell us? It shows that the most common flipper length is around 190 mm. Values much below 190 mm are rare, but a range of values above this are possible, with higher values becoming less frequent. These binned data are telling us a lot about the sample distribution of `flipper_length_mm`.

It is still difficult to perceive distributional information when presented as a series of numbers. What we need is a visualisation based on the above numbers. This is what a **histogram** provides. Histograms summarise the sample distribution of a variable by showing the counts of binned data as a series of bars. The position and width of each bar correspond to an interval and the height shows the count. Here is a histogram based on the binned data we made:

```{r, echo=FALSE}
ggplot(penguins, aes(x = flipper_length_mm)) + geom_histogram(breaks = seq(170, 235, by = 5))
```

This gives a clear summary of the sample distribution of body mass. It reveals:

1.  the most common values, which are round about 190 mm;
2.  the range of the data, which is about 65 mm; and
3.  the shape of the distribution, which appears to be **bimodal**---it has two peaks.

That last observation wasn't immediately obvious when looking at the raw counts in the binned summary.

We used **ggplot2** to make that histogram. We could have used **dplyr** to make the binned data set and then used this with **ggplot2** to construct the plot manually. However, there is a much easier way to make a histogram. We'll demonstrate this now. Rather than do it using one **ggplot2** expression, we will break the process up to demonstrate how it works.

The first step uses the `ggplot` function with `aes` to set up the default data and aesthetic mapping:

```{r}
# create graphical object with data and aesthetic mapping
plot_hist <- ggplot(penguins, aes(x = flipper_length_mm))
```

This is no different than the scatter plot example we stepped through. The only difference is that a histogram requires only one aesthetic mapping. We supplied the argument `x = flipper_length_mm` to `aes` because we want to display the map intervals associated with `flipper_length_mm` to the x-axis. We don't need to supply an aesthetic mapping for the y-axis because **ggplot2** is going to handle this for us.

The second step adds a layer to the `plot_hist` object. We need to find the right `geom_XX` function to do this. This is called `geom_histogram`:

```{r}
# use geom_histogram to add histogram layer
plot_hist <- plot_hist + geom_histogram()
```

It is instructive to look at the summary of the **ggplot2** object at this stage:

```{r}
# print summary of graphical object
summary(plot_hist)
```

Look at the text below the `----`. This shows that `geom_histogram` adds a 'stat' to the layer: `stat_bin`. This means **ggplot2** is going to take the raw `flipper_length_mm` data and bin it for us. Remember---the stat facility of **ggplot2** is its mechanism for creating summaries of data from the raw inputs. `stat_bin` deals with the binning process for us.

Everything we need to plot a histogram is now set up. Here is the resulting plot:

```{r}
# show the histogram
print(plot_hist)
```

This is not quite the same as the example we saw above because it uses different bins. We can set the properties of the `geom_histogram` to tweak this kind of thing---the `binwidth` argument adjusts the width of the bins used. Let's construct the histogram again with 5 mm wide bins, as well as adjust the colour scheme and axis labels a bit:

```{r}
# display flipper length histogram
ggplot(penguins, aes(x = flipper_length_mm)) + 
  # use geom_histogram to add histogram layer + customisation
  geom_histogram(binwidth = 5, fill = "steelblue", colour="darkgrey") + 
  # set the the x- and y-axis labels
  labs(x = "Flipper length (mm)", y = "Count")
```

Notice that the effect of increasing the bin width is to 'smooth' the histogram, i.e. this version looks less jagged than the last. Whether or not that colour scheme is an improvement is a matter of taste. Mostly we wanted to demonstrate how the `fill` and `colour` arguments change the output--- `fill` sets the fill colour of the bar and `colour` deals with the line around each bar.

::: {.infobox .warning data-latex="{warning}"}
#### Choose your own bin widths {.unnumbered}

It is good practice to experiment with `binwidth` because the value selected by **ggplot2** is seldom optimal. Finding a 'good' value is as much art as science. One way of doing this is to start with a value that is probably too large and then refine it down. For example, we could work out the approximate range of the data by eye and then use 1/10^th^ of that range as an initial value for `binwidth`. Unless the data set is very small, this will result in a histogram that aggregates too much information. Once we have that rough starting value, we then reduce `binwidth` in small steps, re-rendering the plot each time until we end up with something that summarises the distribution well.
:::

We can use pretty much the same R code to produce a histogram summarising the body mass sample distribution:

```{r}
# display body mass histogram
ggplot(penguins, aes(x = body_mass_g)) + 
  # use geom_histogram to add histogram layer + customisation
  geom_histogram(binwidth = 250, fill = "steelblue", colour="darkgrey") +
  # set the the x- and y-axis labels
  labs(x = "Flipper length (mm)", y = "Count")
```

The only things that changed in this example were the aesthetic mapping and the bin width, which we set to 250 to reflect the range of observed body masses. The histogram reveals that the most common body mass (the mode) is around 3500 grams and the range of masses is also about 3500 grams. It also shows that the distribution is **skewed**, i.e. it is asymmetric. We can say even more---the distribution is right-skewed (a.k.a. positive-skewed) because the tail on the right gets further away from the mode than the left tail.

We have to choose those bin widths carefully. Body mass is measured to the nearest 25 grams. This means we should choose a bin width that is some multiple of 25 to produce a meaningful histogram. Look what happens if we set the bin width to 25:

```{r}
# display body mass histogram
ggplot(penguins, aes(x = body_mass_g)) + 
  # use geom_histogram to add histogram layer + customisation -- choose 
  # a sensible bin width
  geom_histogram(binwidth = 25, fill = "steelblue", colour="darkgrey") + 
  # set the the x- and y-axis labels
  labs(x = "Flipper length (mm)", y = "Count")
```

We end up with gaps in the histogram because many bins do not include any data. The take-home message is that we have to pay attention to our data to produce meaningful summaries.

### Dot plots

Histograms are good for visualising sample distributions when we have a reasonably large sample size, e.g. \>100 observations. They are less effective when a sample is small. In this 'small data' situation, it is better to use something called a **dot plot**[^univariate_eda-1].

[^univariate_eda-1]: Not to be confused with the 'Cleveland dot plot'. A standard dot plot summarises a sample distribution. The Cleveland dot plot is something quite different, which summarises the frequencies of a categorical variable. It's meant to serve as a simple alternative to bar charts and pie charts.

```{r, eval=TRUE, echo=FALSE}
# get a small subset of penguins
penguins_small <- penguins %>% 
  filter(species == "Adelie", sex == "female", year == 2007)
```

We have created subset of `penguins` that only retains female Adelie penguin observations from 2007, called `penguins_small`. This is considerably smaller than the original:

```{r, echo=FALSE}
# look at the subset of penguins
penguins_small
```

This is about the right size for using a dot plot. Constructing a dot plot with **ggplot2** is easy once you understand the histogram case. The code is are very similar:

```{r}
# display body mass dot plot
ggplot(penguins_small, aes(x = body_mass_g)) + 
  # use geom_dotplot to add layer + choose a sensible bin width!
  geom_dotplot(binwidth = 100) +
  # set the the x-axis label and remove y-axis label
  labs(x = "Body mass (g)", y = "") +
  # remove y-axis guides
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

Each observation in the data adds one dot, and dots that fall into the same bin are stacked up on top of one another. Notice the bins are not evenly spaced, however. The resulting plot displays the same information as a histogram but tends to be more informative when there are few observations.

The `theme` part of that example is entirely optional.It removes the grid lines and y-axis labels because these are not meaningful when the number of stacked dots represents ' height'. It is unfortunate that **ggplot2** displays them. But then, no package is perfect.

## Exploring categorical variables {#explore-categorical}

We will work with the penguin species (`species`) and observation year (`year`) variables to demonstrate how to explore categorical variables.

### What kind of categorical variable?

Exploring categorical variables is generally simpler than the numeric case. The simplest thing we can do is examine the set of categories. For example, we can print these by extracting a variable with the `$` operator and using the `unique` function:

**species** ---

```{r}
unique(penguins$species)
```

**observation year** ---

```{r}
unique(penguins$year)
```

What kind of variable is `species?` This is obviously categorical---there are no numeric values associated with that variable. But is it ordinal or nominal? It is safe to assume that `species` should be treated as a nominal variable. There is no reasonable way to order the three species---i.e. a statement such as 'Adelie \> Gentoo' is nonsense.

What about `year`? That's a bit more tricky. If we were interested in how some aspect of the data changes over time, we might consider treating year as a numeric variable, or perhaps, as an ordinal categorical variable. Alternatively, if the question is simply, 'do the data vary from one year to the next' without any concern for trends, it's reasonable to treat year as a nominal categorical variable.

This once again illustrates an important idea---the classification of a variable sometimes depends on the analysis goals. The decision matters because it influences how we choose to summarise a variable, how we interpret its relationship with other variables, and whether a specific statistical model is appropriate for our data or not.

For now, let's assume it's fine to treat `year` as a nominal variable.

### Bar plots

We can't say anything about the sample distribution of a categorical variable by just looking at the possible categories. We need to explore the relative frequency of those categories. There is no need to carry out anything as complicated as binning. Instead, we simply count up the number of times each category occurs in the data. The **dplyr** `count` verb will does this:

```{r}
penguins %>% count(species)
```

We saw this result earlier in the book---the most common species is Adelie, followed by the Gentoo and Chinstrap. Gentoo is roughly as common as Adelie, whereas Chinstrap is about half as frequent.

These numbers lead directly to an informative visualisation. The usual graphical tool for summarising categorical variables is the **bar chart** (a.k.a. bar plot or bar graph). A bar chart presents summaries of grouped data with rectangular bars. The lengths of each bar is proportional to the value it represents. When summarising categorical variables, the bar lengths show the raw counts or proportions of each category.

Constructing a bar graph to display category counts is easy with **ggplot2**. We'll do this for the `species` variable. As always, we start by using the `ggplot` function to construct a graphical object containing the default data and aesthetic mapping.

```{r}
# create graphical object with data and aesthetic mapping
plot_bar <- ggplot(penguins, aes(x = species)) 
```

We've called the object `plot_bar` for the obvious reason. Notice that we only need to define one aesthetic mapping---we mapped `species` to the x-axis. This will produce a bar plot with vertical bars.

The next step is to add a layer using one of the `geom_XX` functions. There are two functions we can use to create bar charts in **ggplot2**: `geom_bar` and `geom_col`. By default, `geom_bar` uses the stat facility to count the number of observations in each category, whereas `geom_col` plots the actual numbers we provide.

We want to show the counts of each `species` category, meaning we should use `geom_bar`:

```{r}
# use geom_bar to add bar plot layer
plot_bar <- plot_bar + geom_bar()
```

Now examine the summary of the **ggplot2** object:

```{r}
# print summary of graphical object
summary(plot_bar) 
```

Look at the layer information below `----`. The `geom_bar` function uses `stat_count`. This means **ggplot2** summed the number of observations associated with each category of `species`. Counting a categorical variable is analogous to binning a numeric variable. The only difference is that there is no need to specify bin widths because `species` is categorical. Here's the resulting figure:

```{r, fig.height=4, fig.width=2.2}
# show the bar plot
print(plot_bar)
```

This is exactly the same information produced by the `count` function, only now it's presented in graphical form. We can customise this bar graph with functions like `labs` and by setting various properties inside `geom_bar`. For example:

```{r, fig.height=4, fig.width=2.2}
# display bar plot of species counts
ggplot(penguins, aes(x = species)) + 
  # use geom_bar to add bar plot layer of **counts**
  geom_bar(fill = "rosybrown4", width = 0.7) + 
  # set the the x- and y-axis labels
  labs(x = "Species", y = "Number of Observations")
```

The only new thing here is that we used the `width` argument of `geom_bar` to make the bars a little narrower than the default. Notice that we use `fill` (not `colour`) to change the colour of the bars.

What might we like to change about this plot? The different species appear in alphabetical order. There's nothing particularly meaningful about that order. Perhaps it would make more sense to show them in frequency order from least to most common?

We need to customise the scale associated with the 'x' aesthetic to achieve this. We can start by making a short character vector containing the species names, listing them in the order they need to be shown:

```{r}
# character vector of species names **in required order**
ords <- c("Chinstrap", "Gentoo", "Adelie")
```

Keep an eye on the spelling---R is not forgiving of spelling errors. We can use this with the `limits` argument of the `scale_x_discrete` function to adjust the ordering:

```{r, fig.height=4, fig.width=2.2}
# display bar plot of species counts
ggplot(penguins, aes(x = species)) + 
  # use geom_bar to add bar plot layer of **counts**
  geom_bar(fill = "rosybrown4", width = 0.7) + 
  # customise order of x-axis (limits)
  scale_x_discrete(limits = ords) +
  # set the the x- and y-axis labels
  labs(x = "Species", y = "Number of Observations")
```

We had to use one of the `scale_x_YY` functions to change the way the 'x' aesthetic appears. We use `scale_x_discrete` because 'discrete' is **ggplot2**-speak for 'categorical'.

Is there anything else we might want to change? Sometimes it is clearer to plot horizontal bars, particularly if the categorical axis labels are all bunched together (not really the case here). Flipping the x and y axes makes a horizontal bar chart, which we can achieve with the `coord_flip` function (this is new):

```{r, fig.height=1.8, fig.width=5}
# display bar plot of species counts
ggplot(penguins, aes(x = species)) + 
  # use geom_bar to add bar plot layer of **counts**
  geom_bar(fill = "rosybrown4", width = 0.7) + 
  # customise order of x-axis (limits)
  scale_x_discrete(limits = ords) +
  # set the the x- and y-axis labels
  labs(x = "Species", y = "Number of Observations") +
  # flip x- and y-axes
  coord_flip()
```

We use very similar R code to produce a bar plot summarising the annual counts, i.e. those associated with the `year` categories:

```{r, fig.height=1.8, fig.width=5}
# display bar plot of year counts
ggplot(penguins, aes(x = year)) + 
  # use geom_bar to add bar plot layer of **counts**
  geom_bar(fill = "steelblue", width = 0.7) + 
  # set the the x- and y-axis labels
  labs(x = "Year", y = "Number of Observations") +
  # flip x- and y-axes
  coord_flip()
```

To make this, we changed the aesthetic mapping, colour and the labels, and dropped the `scale_x_discrete` part because the default ordering is acceptable. The resulting plot shows that there are roughly equal numbers of penguins sampled in each year.

There is one small irritation. Look at the year axis labels and guides. These include non-existent 'half years' (e.g. 2007.5) rather than being limited to only the observed years (2007, 2008 and 2009). This has happened because **ggplot2** doesn't know that `year` is meant to be treated as a categorical variable. It sees a number, and so it constructs a continuous numeric axis for `year`.

We need to force **ggplot2** to treat `year` as a categorical 'thing' to avoid this behaviour. There is more than one way to do this. The simplest is to convert `year` into a character vector. The `as.character` function will do this. What's more, **ggplot2** will allow us to do this when we set up the aesthetic mappings within `aes`:

```{r, fig.height=1.8, fig.width=5}
# display bar plot of year counts
ggplot(penguins,
       # convert year to character when setting up mapping
       aes(x = as.character(year))) + 
  # use geom_bar to add bar plot layer of **counts**
  geom_bar(fill = "steelblue", width = 0.7) + 
  # set the the x- and y-axis labels
  labs(x = "Year", y = "Number of Observations") +
  # flip x- and y-axes
  coord_flip()
```

That's better! This may seem like a bit of an edge case but it comes up often enough to be worth highlighting. More generally, this example demonstrates that our conception of a variable needs to matches the way it is interpreted by **ggplot2** to arrive at a sensible plot.

```{r, echo=FALSE}
opts_chunk$set(warning=TRUE)
```
