```{r, echo=FALSE}
opts_chunk$set(warning=FALSE)
```

# Doing more with **ggplot2**

We have now seen a range of different ways to visualise and explore data using **ggplot2**. This last chapter will cover a few bits of miscellanea that don't particularly fit anywhere else. That doesn't mean the ideas aren't important!

## Comparing descriptive statistics {#summaries}

So far, we have focussed on displaying the raw data (e.g. scatter plots) or a distributional summary (e.g. box plots). What other types of quantities might we need to visualise? Descriptive statistics such as the sample mean are one possibility. These often feature in data analysis when 'comparing groups'.

We need to know how to construct plots that display such summaries. Let's start with a simple question: how does the (arithmetic) mean body mass varies by sex and across penguin species? One option is to produce a bar plot for which the lengths of bars represent the mean body mass in each category.

There are two different routes to produce this with **ggplot2**. One way to build such a plot is to break the problem into two steps. In the first step, we calculate whatever it is we want to display, i.e. the species- and sex-specific mean body mass. **dplyr** is usually the best tool to use for this first step:

```{r}
penguins_summary <- penguins %>% 
  group_by(species, sex) %>% 
  summarise(mean_mass = mean(body_mass_g)) %>% 
  na.omit()
```

We used `group_by` and `summarise` to calculate the set of means, which we decided to call `mean_mass`. Notice that we used `na.omit` to exclude the `NA` categories that arise when sex is unknown. The resulting `penguins_summary` object is a small 6 row by 3 column tibble:

```{r}
penguins_summary
```

The second step uses `penguins_summary` with **ggplot2** to create the required bar plot:

```{r, fig.width=4}
ggplot(penguins_summary, 
       aes(x = species, fill = sex, y = mean_mass)) +
  geom_col(position = "dodge") + 
  labs(x = "Species", y = "Mean body mass (g)", fill = "Sex")
```

Two points about this are worth noting:

1.  We used `geom_col` instead of `geom_bar`. `geom_bar` counts the observations in each category by default. Using `geom_col` tells **ggplot2** to plot the information in `mean_mass` 'as is'.
2.  We paid close attention to the names of things. The plotting data lives in `penguins_summary` which means the y aesthetic must be associated with `mean_mass` rather than `body_mass_g`.

It is possible to produce this plot with less code by using **ggplot2**'s built-in 'stat' facility (see box). We recommend the long-winded way when starting out with R because it makes it a bit easier to fix mistakes---we can check whether the right information is in a summary data set before plotting it.

::: {.infobox .information data-latex="{information}"}
#### Using **ggplot2** 'stats' {.unnumbered}

A second way to arrive at the above bar plot is by using the built-in **ggplot2** `stat` facility:

```{r, fig.width=4}
ggplot(na.omit(penguins), aes(x = species, fill = sex, y = body_mass_g)) +
  geom_bar(stat = "summary", fun.y = mean, position = "dodge") +
  labs(x = "Species", y = "Mean body mass (g)", fill = "Sex")
```

This involves a couple of new tricks. When we add a layer using `geom_bar` we set two additional arguments:

-   `stat = "summary"` tells **ggplot2** not to plot the raw values of the y aesthetic mapping, but instead, to calculate and then plot a summary of the 'y' variable.
-   `fun.y = mean` tells **ggplot2** how to summarise the 'y' variable. The function on the right-hand side can be any function that takes a vector of values and returns a single number. We want the `mean` function.
:::

### Error bars

A descriptive statistic like the mean isn't much use on its own. To properly interpret means we also need to know something about their uncertainty. There are many ways to quantify the uncertainty of an estimate. Whichever one we are using, displaying the uncertainty typically involves adding **error bars** to a plot.

We'll demonstrate how to do this by extending the previous example to show the mean and standard error of body mass for each species.

::: {.infobox .information data-latex="{information}"}
#### Er.... standard error? {.unnumbered}

The standard error is a measure of how precise an estimate like the sample mean is. A small standard error indicates that we can have more confidence the estimate reflects the 'true' value. The standard error of the mean can be calculated from a well-known formula: $$
\text{Standard Error} = \frac{\text{Standard Deviation}}{\sqrt{(\text{Sample Size})}} = \frac{\sigma}{\sqrt{n}}
$$ It is hard to give an explanation of that formula without knowing a bit of statistical theory---it is just one of those things we have to learn. In terms of code that works with **dplyr**, that calculation looks like this:

```{r, eval=FALSE}
standard_error = sd(x) / sqrt(n())
```

This assumes the numeric variable we are working with is called `x`, and we want the result to be called `standard_error`. Notice this uses a special **dplyr** function called `n` to find the sample size.
:::

Once again, constructing the plot is a two-step process. Start by calculating the means and standard error of body mass for each species and sex combination:

```{r}
penguins_summary <- penguins %>% 
  group_by(species, sex) %>% 
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    ster_mass = sd(body_mass_g, na.rm = TRUE) / sqrt(n())
  ) %>% 
  na.omit()
```

We use **dplyr** to calculate the means and standard errors of each species' body mass, which we called `mean_mass` and `ster_mass`, respectively. Again, we used `na.omit` to exclude the `NA` categories cases. The new `penguins_summary` object is a 6 row by 4 column tibble:

```{r}
penguins_summary
```

Next we use `penguins_summary` to make the plot. We can use `geom_col` to show the means as a bar plot, but what about the error bars? Use the unsurprisingly named `geom_errorbar` to add those! Here is the code:

```{r, fig.width=4}
ggplot(penguins_summary, 
       aes(x = species, fill = sex, y = mean_mass)) + 
  # Plot the means
  geom_col(position = "dodge") + 
  # Add the error bars
  geom_errorbar(aes(ymin = mean_mass - ster_mass, 
                    ymax = mean_mass + ster_mass), 
                width = 0.2, position = position_dodge(0.9)) + 
  # Use a more professional theme 
  theme_classic(base_size = 12) + 
  # Change the axes labels
  labs(x = "Species", y = "Mean body mass (g)", fill = "Sex")
```

The only new things appear in the `geom_errorbar` part:

-   The `ymin` and `ymax` arguments of the `geom_errorbar` function give the lower and upper limits of error bars. Here, we have plotted the mean +/- 1 standard error, i.e. each error bar is two standard errors long.
-   Use the `width` argument to set the width of the 'hat' on each error bar. A value of zero gets rid of it altogether.
-   The position has to be set in two places: `geom_col` and `geom_errorbar`. For a bar plot, we use the not-at-all-intuitive `position = position_dodge(0.9)` to locate the error bars at the centre of each bar.

::: {.infobox .warning data-latex="{warning}"}
#### Warning! {.unnumbered}

Data visualisations can include many different kinds of 'error bar'--- standard deviation, standard error, 95% confidence intervals, etc. When we include any error bars on a plot, we must remember to state which kind was used in the figure legend. Otherwise, readers will have no way of knowing how to interpret the plot.
:::

<!-- Showing data and descriptive statistics--dropped as too much  -->

```{r, eval=FALSE, include=FALSE}
# mean bill length and depth by sex-species combo
penguins_summary <- penguins %>% 
  group_by(species, sex) %>% 
  summarise(
    mean_bill_len = mean(bill_length_mm, na.rm=TRUE), 
    mean_bill_dep = mean(bill_depth_mm,  na.rm=TRUE)
  ) 
# scatter plot of raw data and means
ggplot(penguins, aes(colour = species, shape = sex))  + 
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm),
             alpha = 0.5) +
  geom_point(aes(x = mean_bill_len, y = mean_bill_dep), 
             size = 4, data = penguins_summary)
```

### Alternatives to bar plots

Do we have to use a bar plot to display means? Certainly not. In some ways bar plots are a bit old fashioned. People often prefer to show a set of means as points with error bars. This kind of visualisation is very easy to make now that we already have working bar plot code:

```{r, fig.width=4}
ggplot(penguins_summary, 
       aes(x = species, colour = sex, y = mean_mass)) + 
  # Plot the means
  geom_point(position = position_dodge(0.5)) + 
  # Add the error bars
  geom_errorbar(aes(ymin = mean_mass - ster_mass, 
                    ymax = mean_mass + ster_mass), 
                width = 0.2, position = position_dodge(0.5)) + 
  # Use a more professional theme 
  theme_classic(base_size = 12) + 
  # Change the axes labels
  labs(x = "Species", y = "Mean body mass (g)", fill = "Sex")
```

We made three small changes to the bar plot code to produce this. We...

-   mapped `sex` to the `colour` aesthetic instead of `fill`,
-   switched from using `geom_col` to `geom_point`, and
-   altered the position adjustments to 'dodge' by a smaller amount.

That's it!

## Adding text annotations

We often need to include text annotations on a plot. For example, we might want to show sample sizes or highlight 'statistically significant' differences. Adding text is no different from adding other kinds of objects to a plot---we use an appropriate geom (e.g. `geom_text`) to include textual information. Let's see how this works by adding sample size labels the bar chart we've been building.

We start by placing the labels we want to use in a data frame (or tibble) alongside any other variables used in aesthetic mappings. This can be distinct from the data set used to construct the plot but its often simplest to put everything together in one place. Let's do that by rebuilding the `penguins_summary` data set to include the mean body mass, standard error and sample size for each group.

```{r}
penguins_summary <- penguins %>% 
  group_by(species, sex) %>% 
  summarise(
    mean_mass = mean(body_mass_g), 
    samp_size = n(),
    ster_mass = sd(body_mass_g, na.rm = TRUE) / sqrt(samp_size)
  ) %>% 
  na.omit()
```

Then we add the text showing the sample size to our plot using the function `geom_text`:

```{r, fig.width=4}
ggplot(penguins_summary, 
       aes(x = species, fill = sex, y = mean_mass)) + 
  # Plot the means
  geom_col(position = "dodge") + 
  # Add the error bars
  geom_errorbar(aes(ymin = mean_mass - ster_mass, 
                    ymax = mean_mass + ster_mass), 
                width = 0.2, position = position_dodge(0.9)) + 
  # Add the text
  geom_text(aes(label = samp_size, y = mean_mass + ster_mass + 200),
            size = 3, colour = "darkgrey",
            position = position_dodge(0.9)) + 
  # Use a more professional theme 
  theme_classic(base_size = 12) + 
  # Change the axes labels
  labs(x = "Species", y = "Mean body mass (g)", fill = "Sex")
```

The `y = mean_mass + ster_mass + 200` aesthetic mapping in `geom_text` positions each number just above the error bar. This works because we placed the textual information in the plotting data set, meaning we can use the x/y variables to set the position of the annotations. We also have to use the correct position adjustment to ensure the annotations are located at each bar's centre. Everything else inside `geom_text` is adjusting the appearance of the actual text.

## Saving plots

Plots can be saved using the **Export** button in the RStudio **Plots** tab. However, plots saved this way often end up with low resolution and very pixelated. The `ggsave` function in **ggplot2** produces much better output and supports a wide range of file types. There's not much to it. Using `ggsave` looks like this:

```{r, eval = FALSE}
ggsave("penguins-plot.pdf", height = 5, width = 5)
```

By default, `ggsave` will save the last plot we made. If we had made the bar chart and then ran that `ggsave` line, R would save a copy of the plot to as a PDF file called [penguins-plot.pdf](images/penguins-plot.pdf). Here is how the function works:

-   The first argument is the path and name of the file we want to create. This example would save the plot to a file called [penguins-plot.pdf]{.ul}. Because we only provide a file name, `ggsave` would save it to the current working directory.

-   `ggsave` supports many different formats (e.g. "pdf", "jpeg", "png", "svg"). This can be set via the `device` argument (e.g. `device = "jpeg"`). Alternatively, if we don't set the format, **ggplot2** will guess it from the file extension.

-   The `width` and `height` arguments specify the plot dimensions. These usually require a bit of experimentation to get right. If we do not specify them, `ggsave` will use the current size of the plotting window.

## Multi-panel plots

We have seen how to use `facet_wrap` and `facet_grid` to produce multi-panel plots. These functions are useful when a plot needs to show the same visualisation for different subsets of data. How do we produce a figure that shows a different visualisation in each panel?

There are many ways to do this using an external add-on package. We'll examine the framework provided by the [**cowplot** package](https://wilkelab.org/cowplot/articles/index.html). Cowplot provides various features that help create high-quality figures, including custom themes and facilities to arrange plots into multi-panel figures.

Constructing a multi-panel plot with **cowplot** happens in two stages. First, make the individual component plots using **ggplot2** as usual, but instead of 'printing' them, assign each graphical object a name. Then use the `plot_grid` function from **cowplot** to construct the multi-panel plot from the stored objects.

For example, assume we want to display at the bill morphology scatter plot and the mean body mass bar chart side-by-side. First make the two plots assign them names:

```{r}
# first plot
plt_a <- 
  ggplot(penguins, 
         aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) + 
  geom_point() + 
  scale_color_manual(values = c("cornflowerblue", "seagreen", "orangered3")) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", colour = "Species") +
  theme_classic(base_size = 12)
# second plot
plt_b <- 
  ggplot(penguins_summary, 
         aes(x = species, fill = sex, y = mean_mass)) + 
  geom_col(position = "dodge") + 
  geom_errorbar(aes(ymin = mean_mass - ster_mass, ymax = mean_mass + ster_mass), 
                width = 0.2, position = position_dodge(0.9)) + 
  labs(x = "Species", y = "Mean body mass (g)", fill = "Sex") +
  theme_classic(base_size = 12)
```

Then use `plot_grid` to produce the two-panel plot:

```{r, fig.width=9.0}
plot_grid(plt_a, plt_b, 
          nrow = 1, rel_widths = c(60, 40), 
          labels = c("a)", "b)"), label_size = 14)
```

Once we've made a multi-panel plot, we can use the `ggsave` function to save it as before. The `plot_grid` function allows the figure to be customised in various ways. For example:

-   `nrow` and `ncol` control the number of plots in each row/column,
-   `labels` sets the labels used for each panel,
-   `label_size` controls the size of the label font, and
-   `rel_widths` and `rel_heights` control how much space each plot gets.

The package website has some [good articles](https://wilkelab.org/cowplot/articles/index.html) that explain how this all works (along with the other **cowplot** facilities).

```{r, echo=FALSE}
opts_chunk$set(warning=TRUE)
```
